Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

### 2021-02-16 doc/SA-Hermaphrodites-wDemography.tex
Edits plus discussion
Edited main text, added first draft of discussion, moved some material to appendix

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
\usepackage[colorlinks=true, allcolors=black]{hyperref}

@@

% Definitions
\def\mathbi#1{\textbf{\em #1}}
\def\mbf#1{\mathbf{#1}}
\def\mbb#1{\mathbb{#1}}
\def\mcal#1{\mathcal{#1}}
\newcommand{\bo}[1]{{\bf #1}}
\newcommand{\tr}{{\mbox{\tiny \sf T}}}
\newcommand{\bm}[1]{\mbox{\boldmath $#1$}}
\newcommand{\om}{\omega}
\DeclareRobustCommand{\firstsecond}[2]{#2}
 \def\linenumberfont{\normalfont\scriptsize}
 \newcommand{\kron}{\otimes}
%===========================================
% Basic info from Journal

@@

%\section*{Outline}

@@

%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section*{Abstract}
%%%%%%%%%%%%%%%%%%%%
\ldots

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%

 Whether populations persist or go extinct depends heavily on their ability to adapt to the environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, environmental shifts \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, or genetic constraints \citep{ConnallonHall2018}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

@@

%Conflicting selection and gene-flow between different classes of individuals, or genetic trade-offs between fitness components, can impose constraints on adaptation with particularly interesting evolutionary consequences \citep{CharlesworthHughes2000, ConnallonHall2018}. 
Fitness trade-offs constraining adaptation have particularly interesting evolutionary consequences. On the one hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase the population's overall extinction risk \citep{holman2013consequences, harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which influences the population's capactiy for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the contribution of genetic constraints to the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Classical population genetic models consider relative fitnesses of genotypes, and therefore do not consider the population dynamical consequences of evolutionary change. To take the population dynamical consequences into account, a model is required that links changes in genotype distributions to population dynamics.  The demographic processes of birth and death drive both changes in genotype distributions and population dynamics, and therefore provide a fundamental link between evolution and ecology \citep{metcalf2007evolutionary}.

In this paper, we use an evolutionary demographic model to study when the parameter conditions predicted to be polymorphic by population genetics models are also demographically viable. We apply the model to study the maintenance of polymorphisms by sexually antagonistic selection in both dioecious and hermaproditic species. A major strength of the model is that we can use empirically estimated demographic rates to parameterize the model, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

Sexually antagonstic selection, due to genetic trade-offs between male and female fitness, are a common feature of sexually reproducing populations, and are thought to contribute not only to the maintenance of genetic variation, but also the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, "sexually antagonistic" selection (abbreviated SA hereafter) can arise whereby beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

The evolutionary dynamics of SA alleles in hermaphrodites differ from those in dioecious populations because individuals may reproduce through a combination of self- and outcross fertilization \citep{Goodwillie2005, Igic2006, JarneAuld2006}. Such mixed-mating systems have important theoretical consequences for the maintenance of SA polymorphisms in hermaphrodite populations. Self-fertilization is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014}; but see \citealt{Tazzyman2015}. Yet, other processes, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, or spatial heterogeneity and the complexity of the life-cycle \citep{Olito-etal-2018,ConnallonSharmaOlito2019} can expand the parameter space for SA polymorphism. Overall, both theoretical predictions and current empirical data suggest there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.
% \bigskip

@@

In this article, we develop a stage-structured mendelian matrix model that accommodates selection through male and female sex-functions in populations of simultaneous hermaphrodites with mixed mating systems. Under obligate outcrossing, the model becomes mathematically identical to the two-sex model of \citet{deVriesCaswell2019b}. Using the model, we jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus our analyses on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the population is also demographically viable (i.e., has a positive intrinsic growth rate). The model also allows for the effects of inbreeding depression to manifest in different life-history stages, enabling us to explore the fitness consequences of early- vs.~late-acting inbreeding depression on the model predictions. Finally, we give an illustrative example of the potential for SA polymorphism in real populations by parameterizing our model with empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

A key finding of our study is that when the intrinsic population growth rate is close to $1$, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space for SA polymorphism. Furthermore, this demographically viable polymorphic parameter space is often biased towards alleles with stronger selection through the female- than male sex function. Self-fertilization can alleviate the demographic costs of balanced SA polymorphisms under some conditions, however, the concommitant effects of inbreeding depression generally exacerbate them in populations with mixed-mating systems. Despite these demographic pitfalls, our example using {\itshape M.~guttatus} appears to show that demographic rates observed in real populations are capable of sustaining large regions of viable SA polymorphic space. Overall, our findings provide a more nuanced picture of the nature of SA genetic variation that we should expect to find in natural populations, where the fate of SA alleles and the populations harboring them is determined jointly by evolutionary and demographic processes.

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{The Model}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Simultaneous hermaphrodites express both male and female sex functions (i.e., can be both a mother and father), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. In the following any parameters, matrices or vectors relating to male sex function are denoted with a prime.

In the model, individuals are classified jointly by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively). Whether individuals were produced through selfing or outcrossing is included in the individual state description because individuals produced through selfing might experience reduced survival, growth, or maturation rates as a consequence of inbreeding depression. To model inbreeding depression, we therefore keep track of how individuals were produced.

Each genotype is characterized by a matrix of transition probabilities (including survival) and a matrix of reproductive output through male and female function. These matrices can include time variation or nonlinearities reflecting the environment or density dependence, but these will not be modelled here. Each stage contributes offspring to genotypes at the next time step according to transition matrices that are determined by the mating system and stage-structure of the population.

@@

\noindent where $\mbf{n}^{S}_{i}$ and $\mbf{n}^{X}_{i}$ ($i \in \{AA,Aa,aa\}$) are the stage $\times$ genotype distribution vectors of individuals produced by self-fertilization and outcrossing, respectively. The proportional population vector is given by
\begin{linenomath*}
\begin{equation} \label{eq:propPopVec}
	\tilde{\mbf{p}}(t) = \frac{\tilde{\mbf{n}}(t)}{ \| \tilde{\mbf{n}}(t) \|},
\end{equation}
\end{linenomath*}

@@

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution via female and male sex-functions of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime}_{i}$. Deviations from this assumption are straightforward to incorporate but beyond the scope of the current paper. In principle, the model can admit any kind of pleiotropy among demographic traits because the transition and fertility matrices can differ among genotypes in any way. For simplicity, we assume linear, time-invariant demography.

Notational note: The superscript on a matrix in age$\times$stage models conventionally indicates the stage or age of the subsection of the population vector that the matrix is acting on \citep{CaswellEtAl2018}. For example, $\mbf{U}^{S}_{i}$, acts on the vector of individuals of genotype $i$ that were produced through selfing. In the following, we deviate from this convention by also using $S$ and $X$ as superscripts to indicate reproduction of offspring through current selfing or reproduction. That is,  we define the matrices $\mbf{H}^S_{j}(\tilde{\mbf{n}})$ and $\mbf{H}^X_{j}(\tilde{\mbf{n}})$, as the map of the genotypes of the parent in stage $j$ to the genotypes of their offspring produced by selfing and outcrossing, respectively. The $(k, l)$ entry of $\mbf{H}^{X}_{j}$ and $\mbf{H}^{S}_{j}$ is the probability that an offspring of a genotype $l$ mother, of stage $j$, has genotype $k$. Since we assume that how-you-were-produced does not affect how you reproduce, these matrices will act on both $\mbf{n}^{S}$ and $\mbf{n}^{X}$. Likewise, we will write production $\mcal{F}^S$ and $\mcal{F}^X$ to denote offspring production by self-fertilization and outcrossing.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Mating and offspring production under partial selfing}
The inclusion of simultaneous hermaphrodites that can reproduce both through selfing and through outcrossing complicate mating in our model relative to the two-sex stage-structured model of \citet{deVriesCaswell2019b}. Since we assume that how individuals were produced does not affect the relative success of their male gametes, the frequency distribution in the male gametes is obtained simply by summing the vectors of individuals produced through selfing and individuals produced through outcrossing, that is, 
\begin{equation}
 	\mbf{p}^{\prime}=\frac{(\mbf{n}^X + \mbf{n}^S)}{\| (\mbf{n}^X + \mbf{n}^S)\|}.
 \end{equation}

Let $C$ denote the proportion of each individual's ovules that are self-fertilized (the remaining $1 - C$ are outcrossed), and let $\delta$ represent the proportion of self-fertilized zygotes that fail to develop due to inbreeding depression during early development \citep{Charlesworth1987}. In \hl{Appendix A.x}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}). Outcrossing proceeds similarly to a two-sex model of reproduction \citep{deVriesCaswell2019b}. That is, each individual's genotype determines both the number of ovules produced and the paternal "mating success", broadly defined. For hermaphroditic flowering plants, for example, paternal mating success could reflect pollen production, export efficiency, pollen-tube germination and growth rates, among other things \citep{LloydWebb1986, WangBarrett2020, Harder2016}. Depending on their stage and genotype at the SA locus, individuals contribute male gametes to the overall population gamete pool according to the stage $\times$ genotype vector:
\begin{linenomath*}
\begin{equation}
		\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}= 
		\left(
			\begin{array}{ccc}
				\mbf{1}^{\intercal}_{\omega} & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & 0 \\
				0 & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & \mbf{1}^{\intercal}_{\omega} \\
			\end{array} \right)
		\left(
			\begin{array}{ccc}
				\mbf{F}^{\prime}_{AA} & 0 & 0 \\
				0 & \mbf{F}^{\prime}_{Aa} & 0 \\
				0 & 0 & \mbf{F}^{\prime}_{aa} \\
			\end{array} \right)
		\left(
			\begin{array}{c}
				\mbf{p}^{\prime}_{AA} \\
				\mbf{p}^{\prime}_{Aa} \\
				\mbf{p}^{\prime}_{aa} \\
			\end{array} \right).
\end{equation}
\end{linenomath*}

@@

The parent-offspring map for outcrossing is a function of the allele frequencies in the male gamete pool, and is given by 
\begin{linenomath*}
\begin{equation} \label{eq:HX}
	\mbf{H}^X(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} & \frac{1}{2} q^{\prime}_{A} & 0 \\
				q^{\prime}_{a} & \frac{1}{2} & q^{\prime}_{A}  \\
				0 & \frac{1}{2} q^{\prime}_{a} & q^{\prime}_{a} \\
			\end{array} \right)
\end{equation}
\end{linenomath*}

\noindent From left to right, the columns of matrix $\mbf{H}^X(\tilde{\mbf{n}})$ give the genotype distribution of outcrossed offspring produced by a maternal parent of each genotype ($AA$, $Aa$, and $aa$ respectively).

Reproduction via self-fertilization differs from outcrossing. As in previous models of partial selfing, we assume that individuals produce enough male gametes to easily self-fertilize all of their ovules, and that self-fertilization involves little or no selection from external factors relative to outcrossing (e.g., \citealt{Charlesworth1978a,JordanConnallon2014,Olito2017}, but see \citealt{Tazzyman2015}). Under these assumptions, the genotype distributions of selfed offspring are determined entirely by the parental genotype and the probabilities of segregation and fertilization during and after meiosis: 
\begin{linenomath*}
\begin{equation} \label{eq:HS}
	\mbf{H}^S(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				1 & 1/4 & 0 \\
				0 & 1/2 & 0 \\
				0 & 1/4 & 1 \\
			\end{array} \right).
\end{equation}
\end{linenomath*}

\noindent The columns of $\mbf{H}^S(\tilde{\mbf{n}})$ again give the selfed offspring genotype distributions for parental genotypes of $AA$, $Aa$, and $aa$, respectively.

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Population projection}

Using the component matrices described above, the matrix $\tilde{\mbf{A}}[\tilde{\mbf{n}}]$ that projects the eco-evolutionary dynamics is:
\begin{linenomath*}
\begin{equation} \label{eq:Atilde}
	\tilde{\mbf{A}}[\tilde{\mbf{n}}] = 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{U}^S & \mbf{0} \\ \hline
				\mbf{0} & \mcal{U}^X \\
			\end{array} \right)}_{\tilde{\mbf{U}}} + 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) & \mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) \\ \hline
				\mcal{F}^X(\tilde{\mbf{p}}) (1 - C) & \mcal{F}^X(\tilde{\mbf{p}}) (1 - C)\\
			\end{array} \right)}_{\tilde{\mbf{F}}}.
\end{equation}
\end{linenomath*}

We construct $\tilde{\mbf{A}}[\tilde{\mbf{n}}]$ using the vec-permutation approach of \citet{CaswellEtAl2018}. Briefly, we construct a set of block-diagonal matrices: e.g.,
\begin{linenomath*}
\begin{equation} \label{eq:BlkUS}
	\mbb{U}^S = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),
\end{equation}
\end{linenomath*}

>>>>

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}

@@

% Definitions
\def\mathbi#1{\textbf{\em #1}}
\def\mbf#1{\mathbf{#1}}
\def\mbb#1{\mathbb{#1}}
\def\mcal#1{\mathcal{#1}}
\newcommand{\bo}[1]{{\bf #1}}
\newcommand{\tr}{{\mbox{\tiny \sf T}}}
\newcommand{\bm}[1]{\mbox{\boldmath $#1$}}
\newcommand{\om}{\omega}
\DeclareRobustCommand{\firstsecond}[2]{#2}
 \def\linenumberfont{\normalfont\scriptsize}
 \newcommand{\kron}{\otimes}

 %===========================================
% Basic info from Journal

@@

%\section{Outline}

@@

\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section{Abstract}
%%%%%%%%%%%%%%%%%%%%
\ldots

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%

Population persistence or extinction depends heavily on the ability of a species to adapt to the environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, environmental shifts \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, or genetic constraints \citep{ConnallonHall2018}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

@@

%Conflicting selection and gene-flow between different classes of individuals, or genetic trade-offs between fitness components, can impose constraints on adaptation with particularly interesting evolutionary consequences \citep{CharlesworthHughes2000, ConnallonHall2018}. 
Fitness trade-offs constraining adaptation have particularly interesting evolutionary consequences. On the one hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase the population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which influences the population's capactiy for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the contribution of genetic constraints to the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection, due to genetic trade-offs between male and female fitness, is a common feature of sexually reproducing populations, and is thought to contribute not only to the maintenance of genetic variation, but also to the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, "sexually antagonistic" selection (abbreviated SA hereafter) can arise whereby beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

The evolutionary dynamics of sexually antagonistic alleles in hermaphrodites differ from those in dioecious populations because individuals may reproduce through a combination of self- and outcross fertilization \citep{Goodwillie2005, Igic2006, JarneAuld2006}. For example, self-fertilization is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014}; but see \citealt{Tazzyman2015}). Yet, other processes, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, or spatial heterogeneity and the complexity of the life-cycle can expand the parameter space for SA polymorphism \citep{Olito-etal-2018,ConnallonSharmaOlito2019}. Overall, both theoretical predictions and current empirical data suggest there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes, and therefore do not consider the population dynamical consequences of evolutionary change. To take the population dynamical consequences into account, a model is required that links changes in genotype distributions to population dynamics.  \cite{harts2014demography} build such a model using an individual based approach. Their model tracks both the population dynamics and the spread of sexually antagonistic genes as well as dispersal genes in a species with non-overlapping generations. \cite{deVriesCaswell2019b} introduced a model with overlapping generations, intralocus sexual antagonism, and population dynamics but did not perform analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model by \cite{deVriesCaswell2019b} to include both dioecious and hermaphroditic species, and we use this model to study when the parameter conditions predicted to be polymorphic by population genetics models are also demographically viable. A major strength of the model is that we can use empirically estimated demographic rates to parameterize the model, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

% \bigskip

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Model}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Simultaneous hermaphrodites express both male and female sex functions (i.e., can be both a mother and father), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to male sex function with a prime.

Individuals in the model are jointly classified by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively). Whether individuals were produced through selfing or outcrossing is included in the individual state description because individuals produced through selfing might experience reduced survival, growth, or maturation rates as a consequence of inbreeding depression. To model inbreeding depression, we therefore keep track of how individuals were produced. Of course the level of inbreeding depression will not only be a function of whether the current individual is produced through inbreeding but also of how many generations of inbreeding have occurred (see for example, \cite{kelly1999response} and \cite{kelly2007mutation}) but keeping track of the history of inbreeding is beyond the scope of this paper. Therefore we use the commonly made assumption that individuals produced through selfing suffer a fixed level of inbreeding depression (\textbf{which reference(s) would work best here?}).

% Each genotype is characterized by a matrix of transition probabilities (including survival), and a matrix of reproductive output through male and female function. These matrices can include time variation or nonlinearities reflecting the environment or density dependence (see for example, \cite{de2020matrix}), but these will not be modelled here. Each stage contributes offspring to genotypes at the next time step according to transition matrices that are determined by the mating system and stage-structure of the population.

@@

\noindent where $\mbf{n}^{S}_{i}$ and $\mbf{n}^{X}_{i}$ ($i \in \{AA,Aa,aa\}$) are the stage distribution vectors of individuals of genotype $i$ produced by self-fertilization and outcrossing, respectively. The proportional population vector is given by
\begin{linenomath*}
\begin{equation} \label{eq:propPopVec}
	\tilde{\mbf{p}}(t) = \frac{\tilde{\mbf{n}}(t)}{ \| \tilde{\mbf{n}}(t) \|},
\end{equation}
\end{linenomath*}

@@

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution via female and male sex-functions of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$. Deviations from this assumption are straightforward to incorporate but beyond the scope of the current paper.

% Notational note: The superscript on a matrix in age$\times$stage models conventionally indicates the stage or age of the subsection of the population vector that the matrix is acting on (see for example \cite{CaswellEtAl2018}). For example, $\mbf{U}^{S}_{i}$, acts on the vector of individuals of genotype $i$ that were produced through selfing. In the following, we deviate from this convention by also using $S$ and $X$ as superscripts to indicate production of offspring through current selfing or reproduction. That is,  we will write production $\mcal{F}^S$ and $\mcal{F}^X$ to denote offspring production by self-fertilization and outcrossing. Likewise, we define the matrices $\mbf{H}^S_{j}(\tilde{\mbf{n}})$ and $\mbf{H}^X_{j}(\tilde{\mbf{n}})$ as the map of the genotypes of the parent in stage $j$ to the genotypes of their offspring produced by selfing and outcrossing, respectively. The $(k, l)$ entry of $\mbf{H}^{X}_{j}$ and $\mbf{H}^{S}_{j}$ is the probability that an offspring of a genotype $l$ mother, of stage $j$, has genotype $k$. Since we assume that how-you-were-produced does not affect how you reproduce, these matrices will act on both $\mbf{n}^{S}$ and $\mbf{n}^{X}$.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Mating and offspring production under partial selfing}
 Outcrossing in our model proceeds similarly to a two-sex model of reproduction \citep{deVriesCaswell2019b}. That is, each individual's genotype determines both the number of ovules produced and the paternal "mating success", broadly defined. For hermaphroditic flowering plants, for example, paternal mating success could reflect pollen production, export efficiency, pollen-tube germination and growth rates, among other things \citep{LloydWebb1986, WangBarrett2020, Harder2016}.

Since we assume that how individuals were produced does not affect their relative mating success, the frequency in the male mating population is obtained simply by summing the vectors of individuals produced through selfing and through outcrossing, that is, 
\begin{equation}
 	\mbf{p}^{\prime}=\frac{(\mbf{n}^X + \mbf{n}^S)}{\| (\mbf{n}^X + \mbf{n}^S)\|}.
 \end{equation} 
Depending on their stage and genotype at the SA locus, individuals contribute male gametes to the overall population gamete pool according to the following equation
\begin{linenomath*}
\begin{equation}
		\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}= 
		\left(
			\begin{array}{ccc}
				\mbf{1}^{\intercal}_{\omega} & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & 0 \\
				0 & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & \mbf{1}^{\intercal}_{\omega} \\
			\end{array} \right)
		\left(
			\begin{array}{ccc}
				\mbf{F}^{\prime}_{AA} & 0 & 0 \\
				0 & \mbf{F}^{\prime}_{Aa} & 0 \\
				0 & 0 & \mbf{F}^{\prime}_{aa} \\
			\end{array} \right)
		\left(
			\begin{array}{c}
				\mbf{p}^{\prime}_{AA} \\
				\mbf{p}^{\prime}_{Aa} \\
				\mbf{p}^{\prime}_{aa} \\
			\end{array} \right).
\end{equation}
\end{linenomath*}

@@

\textcolor{red}{
I'm doubting whether to move the next paragraph to the appendices as well.  Equation \ref{eq:maleGametePool} is all we strictly need. If we keep the parent-offspring maps, we can explain in the next section how the matrices 
$\mcal{F}^S$ and $\mcal{F}^X$ have the same structure. Would that be helpful? Or would it be easier to just explain in the next section what the structure of $\mcal{F}^S$ and $\mcal{F}^X$ is? }

The parent-offspring map for outcrossing is a function of the allele frequencies in the male gamete pool, and is given by 
\begin{linenomath*}
\begin{equation} \label{eq:HX}
	\mbf{H}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} & \frac{1}{2} q^{\prime}_{A} & 0 \\
				q^{\prime}_{a} & \frac{1}{2} & q^{\prime}_{A}  \\
				0 & \frac{1}{2} q^{\prime}_{a} & q^{\prime}_{a} \\
			\end{array} \right)
\end{equation}
\end{linenomath*}
\noindent From left to right, the columns of matrix $\mbf{H}^X$ give the genotype distribution of outcrossed offspring produced by a maternal parent of each genotype ($AA$, $Aa$, and $aa$ respectively).

Reproduction via self-fertilization differs from outcrossing. As in previous models of partial selfing, we assume that individuals produce enough male gametes to easily self-fertilize all of their ovules, and that self-fertilization involves little or no selection from external factors relative to outcrossing (e.g., \citealt{Charlesworth1978a,JordanConnallon2014,Olito2017}, but see \citealt{Tazzyman2015}). Under these assumptions, the genotype distributions of selfed offspring are determined entirely by the parental genotype and the probabilities of segregation and fertilization during and after meiosis: 
\begin{linenomath*}
\begin{equation} \label{eq:HS}
	\mbf{H}^S(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				1 & 1/4 & 0 \\
				0 & 1/2 & 0 \\
				0 & 1/4 & 1 \\
			\end{array} \right).
\end{equation}
\end{linenomath*}
\noindent The columns of $\mbf{H}^S(\tilde{\mbf{n}})$ again give the selfed offspring genotype distributions for parental genotypes of $AA$, $Aa$, and $aa$, respectively.

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Population projection}
Using the component matrices described above (the survival matrices, $\mathbf{U}^X_i$, $\mathbf{U}^S_i$, the fertility and mating success matrices, $\mathbf{F}_i$, $\mathbf{F}^\prime_i$, and the parent-offspring matrices $\mbf{H}^S(\tilde{\mbf{n}})$ and $\mbf{H}^X$), we construct the population projection matrix $\tilde{\mbf{A}}[\tilde{\mbf{n}}]$ using the vec-permutation approach of \citet{CaswellEtAl2018}, see \hl{Appendix A.1} for the step by step construction of the model. The matrix that projects the eco-evolutionary dynamics is:
\begin{linenomath*}
\begin{equation} \label{eq:Atilde}
	\tilde{\mbf{A}}[\tilde{\mbf{n}}] = 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{U}^S & \mbf{0} \\ \hline
				\mbf{0} & \mcal{U}^X \\
			\end{array} \right)}_{\tilde{\mbf{U}}} + 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) & \mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) \\ \hline
				\mcal{F}^X(\tilde{\mbf{p}}) (1 - C) & \mcal{F}^X(\tilde{\mbf{p}}) (1 - C)\\
			\end{array} \right)}_{\tilde{\mbf{F}}},
\end{equation}
\end{linenomath*}
where $C$ denotes the proportion of each individual's ovules that are self-fertilized (the remaining $1 - C$ are outcrossed), and  $\delta$ represents the proportion of self-fertilized zygotes that fail to develop due to inbreeding depression during early development \citep{Charlesworth1987}. In \hl{Appendix A.x}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}).
